Personal Total Income of Individual in the United States of America in 2023

### 2024-10-26 paper/paper.qmd
Finish EDA in paper and revise the graph position in Data Section

#| echo: false
#| label: fig-figure1
#| warning: false
#| fig-cap: Dwelling Ownership Details

import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns
data = pd.read_parquet('../data/02-analysis_data/cleaned_data.parquet')
plt.figure(figsize=(10, 6))
sns.countplot(x='MORTGAGE', data=data, palette='pastel', edgecolor='black')
plt.xlabel('Mortgage Status', fontsize=12)
plt.ylabel('Count', fontsize=12)
plt.xticks(rotation=30, ha='right')
plt.grid(axis='y', linestyle='--', alpha=0.7)
plt.tight_layout()
plt.show()

@@

#| echo: false
#| label: fig-figure2
#| warning: false
#| fig-cap: Sex Distribution

import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns
data = pd.read_parquet('../data/02-analysis_data/cleaned_data.parquet')
plt.figure(figsize=(10, 6))
sns.countplot(x='SEX', data=data, palette='pastel', edgecolor='black')
plt.xlabel('Sex', fontsize=12)
plt.ylabel('Count', fontsize=12)
plt.xticks(ha='right')
plt.grid(axis='y', linestyle='--', alpha=0.7)
plt.tight_layout()
plt.show()

@@

#| echo: false
#| label: fig-figure3
#| warning: false
#| fig-cap: Age Distribution

@@

#| echo: false
#| label: fig-figure4
#| warning: false
#| fig-cap: Counts of Marital Status

@@

#| echo: false
#| label: fig-figure5
#| warning: false
#| fig-cap: Educational Attainment

@@

#| echo: false
#| label: fig-figure6
#| warning: false
#| fig-cap: School Type

@@

#| echo: false
#| label: tbl-table1
#| warning: false
#| fig-cap: Veteran Status

@@

#| echo: false
#| label: fig-figure7
#| warning: false
#| fig-cap: Total Income Distribution

@@

@@

Note that according to our previous analysis in @sec-eda, the interaction term, `EDUC_SEX_INTERACTION`, was included to capture the combined effect of education level and gender on income. This allowed us to model the non-linear relationship between these two variables and their joint impact on the outcome variable.

>>>>

#| echo: false
#| label: fig-figure1
#| warning: false
#| message: false
#| fig-pos: "H"
#| fig-cap: Mortgage Status Distribution

import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns
data = pd.read_parquet('../data/02-analysis_data/cleaned_data.parquet')
plt.figure(figsize=(5, 3))
sns.countplot(x='MORTGAGE', data=data, palette='pastel', edgecolor='black')
plt.xlabel('Mortgage Status', fontsize=12)
plt.ylabel('Count', fontsize=12)
plt.xticks(rotation=30, ha='right')
plt.grid(axis='y', linestyle='--', alpha=0.7)
plt.tight_layout()
plt.show()

@@

#| echo: false
#| label: fig-figure2
#| warning: false
#| message: false
#| fig-pos: "H"
#| fig-cap: Sex Distribution

import pandas as pd
import matplotlib.pyplot as plt
import seaborn as sns
data = pd.read_parquet('../data/02-analysis_data/cleaned_data.parquet')
plt.figure(figsize=(5, 3))
sns.countplot(x='SEX', data=data, palette='pastel', edgecolor='black')
plt.xlabel('Sex', fontsize=12)
plt.ylabel('Count', fontsize=12)
plt.xticks(ha='right')
plt.grid(axis='y', linestyle='--', alpha=0.7)
plt.tight_layout()
plt.show()

@@

#| echo: false
#| label: fig-figure3
#| warning: false
#| message: false
#| fig-pos: "H"
#| fig-cap: Age Distribution

@@

#| echo: false
#| label: fig-figure4
#| warning: false
#| message: false
#| fig-pos: "H"
#| fig-cap: Counts of Marital Status

@@

#| echo: false
#| label: fig-figure5
#| warning: false
#| message: false
#| fig-pos: "H"
#| fig-cap: Educational Attainment

@@

#| echo: false
#| label: fig-figure6
#| warning: false
#| message: false
#| fig-pos: "H"
#| fig-cap: School Type

@@

#| echo: false
#| label: tbl-table1
#| warning: false
#| message: false
#| fig-pos: "H"
#| fig-cap: Veteran Status

@@

#| echo: false
#| label: fig-figure7
#| warning: false
#| message: false
#| fig-pos: "H"
#| fig-cap: Total Income Distribution

@@

## Income Distribution by Mortgage Status

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../other/graphs/EDAincomebymortgage.png}
    \caption{Income Distribution by Mortgage Status}
    \label{fig:income-mortgage}
\end{figure}

Figure \ref{fig:income-mortgage} illustrates the income distribution for individuals with different mortgage statuses. The highest average income is observed among individuals with a "Mortgage/Trust Deed," followed by those who own their homes outright ("Fully Owned"). Those with a "Signed Purchase Contract" have the lowest average income among the groups. This pattern may reflect a correlation between mortgage status and income level, where individuals with higher incomes are more likely to secure traditional mortgage loans, while those with lower incomes tend towards alternative purchasing arrangements. This could indicate the influence of financial stability and access to credit on mortgage status.

## Correlation Analysis of Age, Education, and Income

\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../other/graphs/EDAcorrelation.png}
    \caption{Correlation Matrix of Age, Education, and Income}
    \label{fig:correlation-matrix}
\end{figure}

Figure \ref{fig:correlation-matrix} shows the correlation matrix for three key variables: age, education level (EDUC_new), and total personal income (INCTOT). The matrix reveals that there is a moderate positive correlation between education level and income (0.29), suggesting that higher education levels are generally associated with higher income. Additionally, a smaller positive correlation exists between age and income (0.14), which might indicate that as individuals grow older, they tend to earn more due to accumulated experience. The correlation between age and education (0.06) is minimal, which is expected as these two variables are less directly related. Overall, this analysis supports the notion that education level plays a more significant role in income variation than age within this dataset.

## Average Income by Education and Gender {#sec-eda3}
\begin{figure}[H]
    \centering
    \includegraphics[width=0.8\textwidth]{../other/graphs/EDAinteraction.png}
    \caption{Average Income by Education and Sex}
    \label{fig:income-education-sex}
\end{figure}

Figure \ref{fig:income-education-sex} presents the average total personal income based on different levels of education for both male and female groups. From the plot, it is observed that individuals with higher levels of education, particularly those who attended "College 5+ Years," tend to earn more on average compared to other groups. However, there are some variations in income within each education level, as indicated by the error bars. Generally, males tend to have higher average incomes across most education levels. This difference suggests a possible gender-based income disparity, even after controlling for education levels. Additionally, the variability in income among education levels highlights that factors beyond educational attainment, such as occupation and industry, may further impact income. Thus, incorporating an interaction term between education and gender in the model is necessary to capture these nuanced effects more accurately.

@@

Note that according to our previous analysis in @sec-eda3, the interaction term, `EDUC_SEX_INTERACTION`, was included to capture the combined effect of education level and gender on income. This allowed us to model the non-linear relationship between these two variables and their joint impact on the outcome variable.
